Higher-Order unification for free!

### 2024-04-14 paper/main.tex
add todo to section 6

We can now derive the implementation for \maybeeta detection:

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or a name applied to the arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $m - n < i \leq m$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term and \testmaybeeta{} $t$ holds,
  then $t \in \maybeeta$.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\noindent
The rule triggers when the input term \elpiIn{flam F} is in
\maybeeta. It compiles \elpiIn{flam F} to \elpiIn{lam F1} but puts the fresh
variable \elpiIn{A} in its place. This variable sees all the names free in
\elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
which relates the variable \elpiIn{A} with \elpiIn{lam F1}.
This link clearly validates \cref{inv:linklhs}.

\begin{corollary}
  The \rhs of any \linketa has exactly one lambda abstraction, hence
  the rule above respects \cref{inv:link-eta-right}.
  \label{cor:rhs-eta-onelamb}
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above is triggered and that
  the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
  If $\testmaybeeta{}~\lambda y.t_{xy}$ holds the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t_{xy}$ does not hold, also
  $\testmaybeeta{}~\lambda x.\lambda y.t_{xy}$ does not hold, contradicting
  the assumption that the rule triggered.
\end{proof}

\paragraph{Decompilation}
Decompilation of the remaining \linketa (i.e. the \linketa that have been
activated) is performed by iterating over them and unifying \lhs and \rhs. Note
that this unification never fails, since \lhs is a flexible term not appearing
in any other \linketa (by \cref{def:progressetadedup}).

@@

\linketa are meant to delay the unification of ``problematic'' terms until
we know for sure if the term has to be $\eta$-contracted or not.

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$, $T'$ is a term not starting with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \progressetaleft, if 1) \lhs is a name, a constant or an
  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux}, $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) \lhs has \elpiIn{lam} as functor. In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{proof}
  The unification is done between the \rhs of two \linketa. Both \rhs has the
  shape $\lambda x.t$, and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
  Therefore, the unification is done between well-behaved terms.
\end{proof}

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-preservation}
  \label{lemma:unif-wellb}
\end{lemma}
\todo{is this W-enforcing?}

@@

\begin{theorem}[Fidelity in \maybeeta]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, such that \noteD{riprendere sec 3.1}
  $\forall t, t\in\subterm{\foUnifPb} \land t \notin \notllambda$ and the 
  map memory is bijective,
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{A map memory is bijective if any higher-order variable is applied
  with the same number of arguments. This problem is addressed in
  \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch]
  \progressetaleft and \progressetadedup activate a \linketa when, in the
  original unification problem, a \maybeeta term is unified with respectively a
  well-behaved term or another \maybeeta term. In both cases, the links trigger
  a unification which succeeds iff the same unification in \Fo{} succeeds,
  guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
  this progression refines a variable to a rigid term and plays no role in 
  \cref{prop:fidelity}.
  % If 1) a \linketa is activated by \progressetaleft, then unification is done
  % between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
  % activation performs a unification which succeeds iff the original problem in
  % \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
  % unification is done between two \maybeeta terms, and again this unification
  % succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
  % \progressetaright, the unification, done between a variable and a term,
  % always succeeds, this is what we expect to guarantee fidelity, since we
  % are essentially removing a hole added by the compilation in the place of a
  % \maybeeta subterm. In all the cases fidelity is respected.
\end{proof}

@@

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
In turn the rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
$X$ is mapped to $B$ and
$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.

@@

We delay at the end of  next section an example of \linketa progression due to
\progressetaright

>>>>

We can now describe the implementation of \maybeeta detection:

\newcommand{\testmaybeeta}{\emph{maybe-eta}\xspace}
\begin{definition}[\testmaybeeta]\label{def:testmaybeeta}
  Given a $\beta$-normal term
  $s = \lambda x_1 \ldots x_n.t$, \testmaybeeta{} $s$ holds if any
  of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or a name applied to the arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $m - n < i \leq m$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{definition}
\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  If $t$ is a $\beta$-normal term
  and
  $t \in \maybeeta$ then \testmaybeeta{} $t$ holds.
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

@@

\noindent
TWhenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
Invariant~\fullref{inv:linklhs} holds for this link, moreover:

\begin{corollary}
  The \rhs of any \linketa has exactly one lambda abstraction, hence
  the rule above respects \fullref{inv:link-eta-right}.
  \label{cor:rhs-eta-onelamb}
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above is applied and that
  the \rhs of the link is $\lambda x.\lambda y.t$, where $x$ and $y$ occur in $t$.
  If $\testmaybeeta{}~\lambda y.t$ holds then the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t$ does not hold, also
  $\testmaybeeta{}~\lambda x.\lambda y.t$ does not hold, contradicting
  the assumption that the rule was applied.
\end{proof}

\paragraph{Decompilation}
The decompilation of a \linketa is performed by unifying
the \lhs with the \rhs. Note that this unification never fails, since \lhs is a
flexible term not appearing in any other \linketa
(by \cref{def:progressetadedup}).

@@

\linketa{}s are meant to delay the unification of ``problematic'' terms until
we know for sure if the term has to be $\eta$-contracted or not.

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \todo{is w sigmaa defined?}
  \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \progressetaleft, if 1) the \lhs is a name, a constant or an
  application, then, $\lambda x.\lhs \appsep x$ is unified with the \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux},
  $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) the \lhs has \elpiIn{lam} as functor.
  \todo{what does as a functor mean?}
  In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{proof}
  We unify the \rhs{} of two \linketa. Both have the
  shape $\lambda x.t$ and by \cref{lemma:unif-eta-aux}, $\wellb(t)$.
\end{proof}

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-enforcing}
  %\label{lemma:unif-wellb}
\end{lemma}

\begin{proof}[Proof sketch]
  By \cref{def:testmaybeeta}.\todo{shall we move the proof here?}
\end{proof}

\begin{lemma}
  The progress of \linketa guarantees \fullref{prop:w-preservation}
  \label{lemma:unif-wellb}
\end{lemma}

@@

\begin{theorem}[Fidelity in \maybeeta]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, if \noteD{riprendere sec 3.1}
  $\forall t, t\in\subterm{\foUnifPb} \Rightarrow t \notin \notllambda$ and if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch]
  \progressetaleft and \progressetadedup activate a \linketa when, in the
  original unification problem, a \maybeeta term is unified with respectively a
  well-behaved term or another \maybeeta term. In both cases, the links trigger
  a unification which succeeds iff the same unification in \Fo{} succeeds,
  \todo{Fix: show that 5.7 an 5.8 compensate for missing eta rule}
  guaranteeing \cref{prop:fidelity}. \progressetaright never fails, in fact,
  this progression refines a variable to a rigid term and plays no role in 
  \cref{prop:fidelity}.
  % If 1) a \linketa is activated by \progressetaleft, then unification is done
  % between a \maybeeta term $t_1$ and a term $t_2$ (with $\wellb(t_2)$). This
  % activation performs a unification which succeeds iff the original problem in
  % \Fo{} succeeds. If 2) a \linketa is activated by \progressetadedup, then the
  % unification is done between two \maybeeta terms, and again this unification
  % succeeds iff it succeeds in \Fo{}. Finally, if 3) a \linketa is activated by
  % \progressetaright, the unification, done between a variable and a term,
  % always succeeds, this is what we expect to guarantee fidelity, since we
  % are essentially removing a hole added by the compilation in the place of a
  % \maybeeta subterm. In all the cases fidelity is respected.
\end{proof}

@@

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
In turn the rule \todo{which rule} runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
$X$ is mapped to $B$ and
$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.
\todo{explain why decompile does so and which eta?}

@@

We give an example for rule
\progressetaright in the next section.
